\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-02 latex/article_vol2.tex
Renamed to dataMaid

%% almost as usual
\author{Anne H. Petersen\\Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne H. Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{cleanR}: maid for cleaning datasets in R} %% without formatting
\Shorttitle{\pkg{cleanR}: maid for cleaning datasets in R} %% a short title (if necessary)

@@

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools that can be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data, but not necessarily knowledge of \R{R}. Thereby, \pkg{cleanR} 
  aids the dialogue between data
  analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{cleanR} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{cleanR} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and die
  within the \proglang{R} console.

  % The \pkg{cleanR} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data cleaning, quality control, \proglang{R}}
\Plainkeywords{data cleaning, quality control, R} %% without formatting
%% at least one keyword must be supplied

@@

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{cleanR} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{cleanR} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{cleanR}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g. distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{cleanR} will help this necessary knowledge sharing.

While \pkg{cleanR}s primary raison d'être is auto-generating data
quality assessment overview documents, we still wish to emphasize that
it is \emph{not} a tool for unsupervised data cleaning. This qualifies
as the second paradigm of \pkg{cleanR}: Data cleaning decisions
should always be made by humans. Therefore, \pkg{cleanR} does not
supply any tools for ``fixing'' errors in the data. However, we do
provide interactive functions that can be used to identify potentially
erroneous entries in a dataset and that can make it easier to solve
data issues, one variable at a time.

This manuscript is structured as follows: First, in Section
\ref{sec:usingcleanR}, we introduce the main representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{cleanR} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{cleanR}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{cleanR} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{cleanR}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{cleanR} can be used to
solve them.

@@

\section{Creating a data cleaning overview} 
\label{sec:usingcleanR}

The \R{clean()} function is the primary workhorse of \pkg{cleanR} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{cleanR}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{cleanR}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm,page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{cleanR_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

\begin{Schunk}
\begin{Sinput}
> library(cleanR)
> data(toyData)
> toydata
\end{Sinput}
\begin{Soutput}
   var1 var2  var3        var4 var5       var6
1   red    1     a -0.65959383    1 Irrelevant
2   red    1     a  0.08671649    2 Irrelevant
3   red    1     a -0.10951326    3 Irrelevant
4   red    2     a  0.08630221    4 Irrelevant
5   red    2     a -1.84311184    5 Irrelevant
6   red    6     b  0.92210680    6 Irrelevant
7   red    6     b  1.01921086    7 Irrelevant
8   red    6     b -0.92428326    8 Irrelevant
9   red  999     c -0.65340163    9 Irrelevant
10  red   NA     c  0.21133941   10 Irrelevant
11 blue    4     c  0.91783009   11 Irrelevant
12 blue   82     .  0.10313983   12 Irrelevant
13 blue   NA        0.16954218   13 Irrelevant
14 <NA>  NaN other  0.41967230   14 Irrelevant
15 <NA>    5 OTHER  0.77143836   15 Irrelevant
\end{Soutput}
\begin{Sinput}
> clean(toyData)
\end{Sinput}
\end{Schunk}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\ 
  \hline \smallskip
  \textbf{\R{summaryFunction}s}  \smallskip \\
  \quad \R{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\ 
  \quad \R{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\ 
  \quad \R{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \smallskip \\
 \textbf{\R{visualFunction}s} \smallskip \\
  \quad \R{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \smallskip \\
 \textbf{\R{checkFunction}s} \smallskip \\
 \quad \R{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\ 
 \quad \R{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\ 
 \quad \R{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\ 
 \quad \R{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
 \quad \R{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\ 
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{cleanR}. The variable
  classes C, F, I, L, B, N, and D, refer to \R{character}, \R{factor},
  \R{integer}, \R{labelled}, \R{logical} (boolean), \R{numeric}, and \R{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
cleaning summary is manipulated, but also its very contents. This is
done through the summarize-visualize-check (SVC) step, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{cleanR} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \R{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{cleanR} to print overview
lists in \proglang{R}. For example, the implemented
\R{summaryFunction}s are:

@@

Thus we can see, for example, that for \R{numeric}, \R{integer}, or \R{Date} 
variables, \pkg{cleanR} provides functions for adding summary information about 
the minimum and maximum values, while all seven variable
classes dealt with in \pkg{cleanR} have functions for central
tendency summaries (i.e., mode or median).

@@

\section{Using cleanR interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more interactively
through the data cleaning process as well. Aside from the \R{clean()} function 
presented above, \pkg{cleanR} also provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{cleanR}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingcleanR}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \R{allCheckFunctions()} call provide
overviews of the available check functions. 
Moving forward, we limit the numeric checks to only identify miscoded 
missing values:

@@

\section{Extending cleanR} 
\label{sec:extending}

\pkg{cleanR} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{cleanR} is fully
customizable and \R{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{cleanR} output file
can be anything --- depending on the arguments given to \R{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously. 
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \R{summaryFunction}s,
\R{visualFunction}s and \R{checkFunction}s are made.

This section consists of two parts. First, we describe how customized \R{summaryFunction}s, 
\R{visualFunction}s and \R{checkFunctions} can be made, including an introduction to a 
few convenient tools available in \pkg{cleanR} for aiding SVC-function construction. 
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a 
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \R{clean()}.

@@

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary-, visual or check function, one needs
to create a new function with a specific structure.  This can be done with 
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \R{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{cleanR}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{cleanR}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{cleanR} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \R{summaryResult()}, changes the class
of the output, thereby making a \R{print()} method available for it.
Note that \code{v} is a vector and that \code{res} should be either a
character string or something that will be printed as one. In other
words, e.g. integers are allowed, but matrices are not. Though a lot
of different things can go into the \code{summaryFunction} template,
we recommend only using it for summarizing the features of a variable,
and leaving tests and checks for the \code{checkFunction}s (presented
below).

>>>>

%% almost as usual
\author{Anne H. Petersen\\Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne H. Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{dataMaid}: your assistant for data cleaning in R} %% without formatting
\Shorttitle{\pkg{dataMaid}: your data cleaning PA in R} %% a short title (if necessary)

@@

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools that can be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data, but not necessarily knowledge of \R{R}. Thereby, \pkg{dataMaid} 
  aids the dialogue between data
  analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and die
  within the \proglang{R} console.

  % The \pkg{dataMaid} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data cleaning, quality control, \proglang{R}}
\Plainkeywords{data cleaning, quality control, R} %% without formatting
%% at least one keyword must be supplied

@@

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{dataMaid} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{dataMaid} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g. distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

While \pkg{dataMaid}s primary raison d'être is auto-generating data
quality assessment overview documents, we still wish to emphasize that
it is \emph{not} a tool for unsupervised data cleaning. This qualifies
as the second paradigm of \pkg{dataMaid}: Data cleaning decisions
should always be made by humans. Therefore, \pkg{dataMaid} does not
supply any tools for ``fixing'' errors in the data. However, we do
provide interactive functions that can be used to identify potentially
erroneous entries in a dataset and that can make it easier to solve
data issues, one variable at a time.

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{dataMaid} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

\section{Creating a data cleaning overview} 
\label{sec:usingdataMaid}

The \R{clean()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{dataMaid}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm,page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

\begin{Schunk}
\begin{Sinput}
> library(dataMaid)
> data(toyData)
> toydata
\end{Sinput}
\begin{Soutput}
   var1 var2  var3        var4 var5       var6
1   red    1     a -0.65959383    1 Irrelevant
2   red    1     a  0.08671649    2 Irrelevant
3   red    1     a -0.10951326    3 Irrelevant
4   red    2     a  0.08630221    4 Irrelevant
5   red    2     a -1.84311184    5 Irrelevant
6   red    6     b  0.92210680    6 Irrelevant
7   red    6     b  1.01921086    7 Irrelevant
8   red    6     b -0.92428326    8 Irrelevant
9   red  999     c -0.65340163    9 Irrelevant
10  red   NA     c  0.21133941   10 Irrelevant
11 blue    4     c  0.91783009   11 Irrelevant
12 blue   82     .  0.10313983   12 Irrelevant
13 blue   NA        0.16954218   13 Irrelevant
14 <NA>  NaN other  0.41967230   14 Irrelevant
15 <NA>    5 OTHER  0.77143836   15 Irrelevant
\end{Soutput}
\begin{Sinput}
> clean(toyData)
\end{Sinput}
\end{Schunk}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\ 
  \hline \smallskip
  \textbf{\R{summaryFunction}s}  \smallskip \\
  \quad \R{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\ 
  \quad \R{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\ 
  \quad \R{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \quad \R{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
  \smallskip \\
 \textbf{\R{visualFunction}s} \smallskip \\
  \quad \R{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \quad \R{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\ 
  \smallskip \\
 \textbf{\R{checkFunction}s} \smallskip \\
 \quad \R{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\ 
 \quad \R{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\ 
 \quad \R{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\ 
 \quad \R{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\ 
 \quad \R{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\ 
 \quad \R{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\ 
 \quad \R{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\ 
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \R{character}, \R{factor},
  \R{integer}, \R{labelled}, \R{logical} (boolean), \R{numeric}, and \R{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
cleaning summary is manipulated, but also its very contents. This is
done through the summarize-visualize-check (SVC) step, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \R{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\R{summaryFunction}s are:

@@

Thus we can see, for example, that for \R{numeric}, \R{integer}, or \R{Date} 
variables, \pkg{dataMaid} provides functions for adding summary information about 
the minimum and maximum values, while all seven variable
classes dealt with in \pkg{dataMaid} have functions for central
tendency summaries (i.e., mode or median).

@@

\section{Using dataMaid interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more interactively
through the data cleaning process as well. Aside from the \R{clean()} function 
presented above, \pkg{dataMaid} also provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{dataMaid}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingdataMaid}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \R{allCheckFunctions()} call provide
overviews of the available check functions. 
Moving forward, we limit the numeric checks to only identify miscoded 
missing values:

@@

\section{Extending dataMaid} 
\label{sec:extending}

\pkg{dataMaid} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{dataMaid} is fully
customizable and \R{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{dataMaid} output file
can be anything --- depending on the arguments given to \R{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously. 
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \R{summaryFunction}s,
\R{visualFunction}s and \R{checkFunction}s are made.

This section consists of two parts. First, we describe how customized \R{summaryFunction}s, 
\R{visualFunction}s and \R{checkFunctions} can be made, including an introduction to a 
few convenient tools available in \pkg{dataMaid} for aiding SVC-function construction. 
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a 
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \R{clean()}.

@@

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary-, visual or check function, one needs
to create a new function with a specific structure.  This can be done with 
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \R{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{dataMaid}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{dataMaid}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{dataMaid} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \R{summaryResult()}, changes the class
of the output, thereby making a \R{print()} method available for it.
Note that \code{v} is a vector and that \code{res} should be either a
character string or something that will be printed as one. In other
words, e.g. integers are allowed, but matrices are not. Though a lot
of different things can go into the \code{summaryFunction} template,
we recommend only using it for summarizing the features of a variable,
and leaving tests and checks for the \code{checkFunction}s (presented
below).

Adhering to the template above is sufficient for using the freshly
made \R{mySummaryFunction()} in \R{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \R{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\R{description} (an explanatory text which will be added to the attributes of
the function), and \R{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call. 
% One comment should be devoted to
% the two attributes of a \R{summaryFunction}. 
If the \R{description} argument is left unspecified, the name of the
function (which in this case is \R{"mySummaryFunction"}) will be filled in and
stored under the \R{description} attribute. What happens if the \R{classes} argument is
not specified depends on the type of \R{mySummaryFunction}. If
\R{mySummaryFunction} is constructed as an \R{S3} generic function with associated
methods, the call to \R{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \R{mySummaryFunction} is not an \R{S3} generic and
\R{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \R{allClasses()} might be useful
for filling out the \R{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"  
[6] "logical"  "numeric"  
\end{Soutput}
\end{Schunk}
% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}
